\BalancedReachability*
\begin{proof}
\end{proof}

\NPHardness*
\begin{proof}
First, we show \NP-hardness of membership to $\BalReach$ via reduction from the 3-SAT problem. 

Consider a 3-CNF formula $\varphi = \bigwedge_{1\leq i \leq k} C_i$. 
We split every variable $x_j$ in $\varphi$ into $x_j^1, \ldots, x_j^m$, where $m$ is the number of occurrences of $x_j$ in $\varphi$.
Let $\varphi^*$ be obtained from $\varphi$ by substituting every (propositional) variable $x_j$ with $x_j^1, \ldots, x_j^m$ such that each (propositional) variable in $\varphi^*$ occurs exactly once.
Observe that $\varphi$ is satisfiable if and only if $\varphi^*$ has a satisfying assignment, which is consistent among copies of each variable, i.e., 
$x_j^1, \ldots, x_j^m$ have the same value assigned. 

We construct a rule with variables $\lhs \to \rhs$ and two \DFA $\aut_1, \aut_2$ over $\set{t,f,\#}$ 
such that variables in $\lhs \to \rhs$ correspond to variables in $\varphi^*$ and for an instantiation of $\lhs \to \rhs$ 
$\aut_1$ checks that it corresponds to a truth assignment, which is consistent among copies of the variable, an
$\aut_2$ checks that the truth assignment satisfies $\varphi^*$.

Let $\lhs$ be the rule of the from $X_1^1 \# X_1^2 \ldots \# X_1^m \# X_2^1 \# \ldots \# X_n^{m'}$, which consists of all word variables corresponding to propositional variables from $\varphi^*$
which are grouped so that first are all (variables corresponding to) copies of $x_1$, next copies of $x_2$ and so on. There is letter $\#$ between any two variables. 
Then, we define $\aut_1$ as in the following diagram, where all transitions that are not depicted here lead to the sink state.
\begin{figure}[h!]
\centering
\begin{tikzpicture}[
    shorten >=1pt, 
    node distance=2cm, 
    on grid, 
    auto,
    inner sep=0.05cm,
    >={Stealth[bend]}
]


  \node[state] (start) {$q_{1,1}$};
  \node[state] (q01) [above right=of start] {$q^f_{1,2}$};
  \node[state] (q02) [right=of q01] {$q^f_{1,3}$};
  \node[state] (q03) [right=of q02] {$q^f_{1,4}$};
  \node[state] (q04) [right=of q03] {$q^f_{1,5}$};
  \node[draw=none] (dots0) [right=of q04] {$\dots$};
  \node[state] (q0m) [right=of dots0] {$q^f_{1,2m}$};

  \node[state] (q11) [below right=of start] {$q^t_{1,2}$};
  \node[state] (q12) [right=of q11] {$q^t_{1,3}$};
  \node[state] (q13) [right=of q12] {$q^t_{1,4}$};
  \node[state] (q14) [right=of q13] {$q^t_{1,5}$};
  \node[draw=none] (dots1) [right=of q14] {$\dots$};
  \node[state] (q1m) [right=of dots1] {$q^t_{1,2m}$};

  \node[state] (q2)  [below right=of q0m] {$q_{2,1}$};



  \path[->]
    (start) edge node [above] {$f$} (q01)
    (q01) edge node [above] {$\#$} (q02)
    (q02) edge node [above] {$f$} (q03)
    (q03) edge node [above] {$\#$} (q04)
    (q04) edge  (dots0)
    (dots0) edge  (q0m)
    (q0m) edge node [above] {$\#$} (q2)

    (start) edge node [above] {$t$} (q11)
    (q11) edge node [above] {$\#$} (q12)
    (q12) edge node [above] {$t$} (q13)
    (q13) edge node [above] {$\#$} (q14)
    (q14) edge  (dots1)
    (dots1) edge  (q1m)
    (q1m) edge node [above] {$\#$} (q2)
    
    
    ;

 
\end{tikzpicture}
%\label{The gadget \DFA $\autB_{Y,x}$, which parially distingushes words of length at most $n$, which do not contain any letter from $Y$ }
\end{figure}
Observe that if each variable in $\lhs$ is instantiated to $t$ or $f$, then $\aut_1$ moves from $q_{1,1}$ to $q_{1, n+1}$ if and only if 
the corresponding truth assignment is consistent among all copies of the variable from $\varphi$.
If there is an instantiation of $\lhs$ such that $\aut_1$ moves from $q_{1,1}$ to $q_{1, n+1}$ over the resulting word, but some of the variables
is not instantiated to $t$ or $f$, then there is (possibly the same) variable instantiated to a word containing $\#$. 
Next, we define $\rhs$ and $\aut_2$ so that  in the instantiation no variable contains $\#$ and if the corresponding instantiation defines a
truth assignment, then it satisfies $\varphi$.

Let $\rhs$ be the list of variables of all clauses $c_1, \ldots, c_k$ of $\varphi^*$, i.e., first are $3$ variables from $c_1$, next $3$ variables from $c_2$ and so on.
Now, we define $\aut_2$, which is a sequential composition of subautomata $\autB_{j}$ corresponding to clauses $c_1, \ldots, c_k$. 
Assume that $c_j  = y_1 \land \neq y_2 \land y_3$ and $Y_1 Y_2 Y_3$ is a subword in $\rhs$ (the order in $\rhs$ is the same).
Then, the \DFA $\autB_{j}$ is illustrated below:
\begin{figure}[h!]
\centering
\begin{tikzpicture}[
    shorten >=1pt, 
    node distance=2cm, 
    on grid, 
    auto,
    inner sep=0.05cm,
    >={Stealth[bend]}
]


  \node[state] (start) {$s_{j,0}$};
  \node[state] (s11) [above right=of start] {$s^t_{j,1}$};
  \node[state] (s12) [right=of s11] {$s^t_{j,2}$};
  \node[state] (s13) [right=of s12] {$s^t_{j,3}$};
  \node[state] (s14) [right=of s13] {$s_{j+1,0}$};
  
  \node[state] (s01) [right=of start] {$s^f_{j,1}$};
  \node[state] (s02) [right=of s01] {$s^f_{1,2}$};
  

  \path[->]
    (start) edge node [above] {$t$} (s11)
    (s11) edge node [above] {$t,f$} (s12)
    (s12) edge node [above] {$t,f$} (s13)
    (s13) edge node [above] {$t,f$} (s14)
    
    (start) edge node [above] {$f$} (s01)
    (s01) edge node [above] {$t$} (s02)
    (s01) edge node [above] {$f$} (s12)
    (s02) edge node [above] {$t$} (s13)


    
    ;

 
\end{tikzpicture}
%\label{The gadget \DFA $\autB_{Y,x}$, which parially distingushes words of length at most $n$, which do not contain any letter from $Y$ }
\end{figure}
All transitions not depicted here lead to the sink state. 
Consider $\aut_2$ to be the \DFA obtained from $\autB_1, \ldots, \autB_k$, where the states $s_{j+1,0}$ is $\aut_{j}$ and $\aut_{j+1}$ are merged into a single state,
Observe that if there is an instantiation of $\rhs$ such that $\aut_2$ moves from $s_{1,0}$ to $s_{k+1,0}$, then this instantiation does not contain letter $\#$ as it leads to the sink state.
Furthermore, if every variable is instantiated to $f$ or $t$
and $\aut_2$ moves from $s_{1,0}$ to $s_{k+1,0}$, then every clause of $\varphi^*$, and hence it has to satisfy $\varphi^*$.

Now, consider an instantiation of $\lhs \to \rhs$ such that over the instance of $\lhs$ the \DFA $\aut_1$ moves from $q_{1,1}$ to $q_{1, n+1}$ and 
over the instance of $\rhs$ the \DFA $\aut_2$ moves from $s_{1,0}$ to $s_{k+1,0}$, then the instantiation encodes a truth assignment that is consistent among all copies of
the same variable from $\varphi$ and satisfies $\varphi^*$. Thus, $\varphi$ is satisfiable.
Conversely, if $\varphi$ is satisfiable, then the satisfying assignment induces an instantiation of $\lhs \to \rhs$ such that 
$\aut_1$ moves from $q_{1,1}$ to $q_{1, n+1}$ and 
$\aut_2$ moves from $s_{1,0}$ to $s_{k+1,0}$.
Therefore, $(q_{1,1}, s_{1,0},q_{1, n+1},s_{k+1,0})$ belongs to $\BalReach$ of the disjoint union of $\aut_1$ and $\aut_2$ 
 if and only if $\varphi$ is satisfiable. 

Now, consider $\aut$ over $\set{t,f,\#,a_1,a_2}$, which consists of the start state $q_0$ and disjoint union of $\aut_1$ and $\aut_2$.
In the state $q_0$ the automaton has self loops over $t,f,\#$ and it moves over $a_1$ to $q_{1,1}$ and over $a_2$ to $s_{1,0}$.
Finally, in all states of $\aut_1$ and $\aut_2$ letters $a_1,a_2$ lead to the sink state.
Consider a rule $a_1 \lhs \to a_2 \rhs$, where  $\lhs \to \rhs$  is defined as above. 
Then, $(q_0, q_{1, n+1},s_{k+1,0})$ belong to $\DeltaRule$ if and only if $\varphi^*$ is reachable.

Finally, consider $\aut$ defined as above, but we mark $q_{1, n+1}$ and the sink state of $\aut_2$ as accepting. 
If there an instantiation of $a_1 \lhs \to a_2 \rhs$ such that the instance of $a_1 \lhs$ is satisfiable and 
$a_2 \rhs$ is not, then in $\aut_1$ the instance of $\lhs$ leads from $q_{1,1}$ to $q_{1, n+1}$, and
in $\aut_2$ the instance of $\rhs$ leads from $s_{1,0}$ to any other state than the sink, and 
hence no variable instance consists $\#$. It follows that each variable is instantiated to $t$ or $f$, and hence
over the instance of $\rhs$, the automaton $\aut_2$ moves from $s_{1,0}$ to either the sink or $s_{k+1,0}$.
Therefore, it has to be $s_{k+1,0}$. 
In summary, if $a_1 \lhs \to a_2 \rhs$ is positively consistent with $\lang(\aut)$, then $\varphi$ is unsatisfiable.
Conversely, one can see that any satisfying assignment for  $\varphi$  can be transformed to the counterexample to positive consistency. 
Thus, positive consistency is \coNP-hard.

Finally, positive and negative consistency are dual and hence negative consistency is \coNP-hard as well.
\end{proof}



>>>>>>> 40b6fcffd9c3a7cbbad600ff38249e026063da6f
\NPHardnessOfClash*
\begin{proof}
Let us fix a parameter $n$. Before the reduction, we construct a gadget that distinguishes all words of length at most $n$ unless they consist one of the designated letters. 
Let $Y \subseteq \Sigma$ and $x \in \Sigma \setminus Y$. Consider the automaton $\autB_{Y,x}$ depicted below:

\begin{figure}[h!]
\centering
\begin{tikzpicture}[
    shorten >=1pt, 
    node distance=2.5cm, 
    on grid, 
    auto,
    >={Stealth[bend]}
]

  % State Definitions
  \node[state] (qn) {$q_n$};
   \node[draw=none] (dots) [right=of qn] {$\dots$};
  \node[state] (q2) [right=of dots] {$q_2$};
  \node[state] (q1) [right=of q2] {$q_1$};
\node[state] (qF) [right=of q1] {$q_F$};

\node[state] (sink) [below=of q2] {$s$};

  % Transition Edges
  \path[->]
    (qn) edge node [above] {$\Sigma\setminus Y$} (dots)
    (dots) edge node [above] {$\Sigma\setminus Y$} (q2)
    (q2)  edge node [above] {$\Sigma\setminus Y$} (q1)
    % The top arc from qn back to q1
    (q1)  edge [bend right=45] node [above] {$\Sigma\setminus (Y \cup \set{x})$} (qn)
    % The outgoing edge xi from q1
%    (q1)  edge [nodes={at={(1, -0.5)}}] node {$x$} ++(1.5, 0);
    (q1)  edge node [above] {$x$} (qF)
    (q1) edge node [left] {$Y$} (sink)
    (q2) edge node [left] {$Y$} (sink)
    (qn) edge node [above] {$Y$} (sink)
    (qF) edge node [left] {$Y$} (sink)
    ;

 \draw[loop,->] (qF)  to node[above] {$\Sigma \setminus Y$} (qF);
 \draw[loop right,->] (sink) to node[right] {$\Sigma$} (sink);

\end{tikzpicture}
%\label{The gadget \DFA $\autB_{Y,x}$, which parially distingushes words of length at most $n$, which do not contain any letter from $Y$ }
\end{figure}

Observe that for all words $w_1, w_2$ of length bounded by $n$, if $f^{\autB_{Y,x}}_{w_1} = f^{\autB_{Y,x}}_{w_2}$, then one of the following holds:
\begin{itemize}
    \item each of the words $w_1, w_2$ contains at least one letter from $Y$; in that case $f_{w_1}, f_{w_2}$ are constant functions equal 
    $s$ for every argument,
    \item the sets positions at which $x$ occurs in $w_1$ and resp. $w_2$ are equal; 
    indeed, for any $w \in (\Sigma \setminus Y)^{\leq n}$ we have
    $f_{w}(q_i) = q_F$ if and only if $w[i] = x$.
\end{itemize}
Now, consider $\autB_Y$ being the disjoint union of automata $\autB_{Y,x}$ over all $x \in \Sigma\setminus Y$.

Observe that for all words $w_1, w_2$ of length bounded by $n$, if $f^{\autB_{Y}}_{w_1} = f^{\autB_{Y}}_{w_2}$, then one of the following holds:
\begin{itemize}
    \item each of the words $w_1, w_2$ contains at least one letter from $Y$; in that case $f_{w_1}, f_{w_2}$ are constant functions equal 
    $s$ for every argument,
    \item words $w_1$ and $w_2$ are equal.
\end{itemize}

Having the automaton $\autB_Y$ we reduce the 3-SAT problem to the minimal collision problem.
Let $\varphi$ be a 3-CNF formula over variables $x_1, \ldots, x_n$.
Consider $\Sigma = \{x_1, \bar{x_1}, \ldots, x_n, \bar{x_n}, a, b \}$.
With each clause $c_j$ we associate the subset $C_j$ of $\Sigma$ consisting of the literals from $c_j$.
Furthermore, for every variable $x_i$ we define the set $X_i = \set{x_i, \bar{x_i}}$.
Now, consider a \DFA $\aut_0$ being the disjoint union of automata $\autB_{C_1}, \ldots, \autB_{C_m}$ over all clauses of $\varphi$ and
automata $\autB_{X_1}, \ldots, \autB_{X_n}$ over all variables of $\varphi$. 
We extend $\aut_0$ to $\aut$ with additional states such that over these states letters from $\Sigma \setminus \set{a,b}$ are self loops, and
letters $a,b$ ensure that every component of the disjoint union is reachable. 
Let $k$ in the instance of the minimal collision problem be equal $n$, the number of variables.

Now, consider a pair of different words $w_1, w_2$ such that $f^{\aut}_{w_1} = f^{\aut}_{w_2}$ and $|w_1|,|w_2| \leq n$.
Then, for every automaton $\autB_{X_1}, \ldots, \autB_{X_n}$, since $w_1 \neq w_2$, each word contains at least one letter from each set $X_i$.
There are $n$ such sets, which are disjoint, and hence $w_1$ and $w_2$ contain exactly one letter from each of the sets $X_i$, i.e., 
$w_1$ and $w_2$ represent consistent variable assignments (they either contain $x_i$ or $\bar{x_i}$).
Furthermore, for every clause $c_j$ each word $w_1, w_2$ contains at least one letter corresponding to some literal from $c_j$.
Therefore, $w_1$ and $w_2$ represent (possibly different) satisfying assignments for $\varphi$.

Conversely, assume that $\varphi$ has a satisfying assignment. 
Consider two words $w_1, w_2$ representing this variable assignment with a different order or literals.
Then, for every component $\autB_{Y,x}$ of $\aut_0$, the functions $f_{w_1}^{\autB_{Y,x}}, f_{w_2}^{\autB_{Y,x}}$ are constant functions equal $s$.
Furthermore, on states in $\aut$, which are not in $\aut_0$, both functions $f_{w_1}^{\aut}, f_{w_2}^{\aut}$ coincide with the identity.
Consequently, we have
(a)~$w_1 \neq w_2$, (b)~$|w_1| = |w_2| = n$, and
(c)~$f_{w_1}^{\aut_0}, f_{w_2}^{\aut_0}$, i.e., 
the constructed instance of the minimal collision problem is positive.
\end{proof}
