We evaluate our framework by expressing natural properties of \VPLs as rewrite rules with variables. 
Our experiments show that learning with advice yields 
a substantial reduction in the number of equivalence queries, runtime reduction and improved accuracy,
demonstrating the practical advantages of the proposed method.    

\subsection{Examples of rewrite rules}
We consider the following rules, which express natural properties of \VPLs.
\begin{itemize}
\item \emph{(Well-matched) commutativity}: $c_1 c_2 X r_2 r_1 \to c_2 c_1 X r_1 r_2$. Consider a \DVPA 
modelling a system consisting of two pushdown components such that $c_1, r_1$ are actions of the first component,
while $c_2, r_2$ are actions of the second. Then, the trace language of the whole system satisfies 
(well-matched) commutativity.
\item \emph{(Well-matched) cancellation}: $c X r \to \epsilon$. Consider a markup language, in which $c$ is a tag opening comment section,
while $r$ is a tag closing it. In such a case, (well-matched) cancellation holds.
\item \emph{(Well-matched) idempotency}: $c c X r r \to c X r$. Consider a markup language, in which $c$ is a tag activating certain mode 
(e.g. bold text) and $r$ is a tag deactivating it. In this setting (well-matched) idempotency holds.
\end{itemize}

% reserch questions

\subsection{Implementation}

\Paragraph{Baseline}
As a baseline for our experiments, we have implemented a polynomial-time active learning algorithm for \DVPA~\cite{mfcs22}.
The algorithm from~\cite{mfcs22} assumes a stronger teacher model than the typical MAT; it allows extended membership queries
that both read and modify the stack content. Moreover, these queries depend on a concrete \DVPA rather than \VPL, and hence
in this framework the automaton is learned rather than the language. This allowed for
 the learning algorithm to work in polynomial time. 

\Paragraph{Equivalence oracle}
We have implemented two types of equivalence test: the exact test based on reachability in the product automaton, and
the probabilistic test based on testing conformance of automata on a large number ($10^5$ and $10^6$) of randomly generated words
of a random length bounded by $100$.
The probabilistic equivalence test is significantly faster than the exact one (from 10x to even 100x faster for 25-state automata).

\Paragraph{The advice mechanism}
We have extended the baseline algorithm with the advice mechanism, which intercepts equivalence queries before
passing them to the equivalence oracle, and performs the consistency check.
The advice is given via rewrite rules with a {single} variable.
We have implemented the exact algorithm presented in the paper, and two independent heuristics. 
The first heuristic is to instantiate a rule with variables to 40--200 ground rules and work with those.
%We select well-matched words to be pairwise maximally different functions from $Q$ to $Q$, using them as candidates for substitution.
The second heuristic is the probabilistic consistency test, which, analogously to the equivalence test, is
approximated by evaluation of a large number of randomly generated words.
 

\subsection{Experiments}

\Paragraph{Benchmarks}
We have generated random benchmark \DVPA with multiple parameters such as the size of the alphabet, 
the size of the stack alphabet, number of states and others.
%
%, density of transitions (the ratio of defined transitions to all possible transitions, where
%undefined transitions lead to the sink state), and the ratio of accepting states among all states. 
%We have observe that the number of states has the main influence on the number of equivalence queries, while
%the size of the alphabet or the size of the stack alphabet have negligable impact on the number of required equivalence queries.
We have also implemented random generation of  \DVPA that are SEVPA and random \DVPA  that are 
consistent with each of advice rules discussed above. For instance, to randomly generate 
a \DVPA consistent with {(well-matched) commutativity}, we generate two random \DVPA over disjoint alphabets, take the product of these automata, where undefined transitions are defined as self-loops. 

We have conducted a series of experiments to address the following research questions:
\begin{enumerate}[Q1]
\item What is the average reduction in the number of equivalence queries avoided using advice? 
\item Does the use of the advice mechanism lead to better running time of the whole learning setup 
(i.e., both the learning algorithm and the equivalence oracle)? 
\item Does the use of the advice mechanism lead to higher reliability in presence of
the probabilistic equivalence oracle?
\end{enumerate}

Research questions Q1, Q2 have been investigated in the \textbf{exact setting}, where the equivalence check as well as consistency check are 
computed based on  reachability, while questions Q2 and Q3 have been investigated in the \emph{probabilistic setting}, where 
the equivalence and consistency checks are approximated.



\Paragraph{The exact setting}
%In this section, we present the reduction in the number of equivalence queries achieved by using SRS,
%as well as its impact on the average execution time required to process a single DVPA in our setup.
The exact equivalence checks dominate the overall computation time. 
When no inconsistency with an SRS is detected,
we need to perform the actual equivalence check, which amounts to two reachability checks.
However, the consistency check is computationally cheaper, as it operates
on smaller automata; the product of the candidate automaton with itself is considerably smaller than the product of
the candidate automaton with the target automaton.
% and hence the reachability check is performed on the products automaton of considerably smaller size.
%In experiments, the alphabet consists of three symbols of each type. For the well-matched commutativity property,
%both automata use such alphabets, and hence the resulting automaton has twice as many letters. 
%The number of stack symbols is set to $5$ for well-matched cancellation and idempotency,
%and to $16$ for well-matched commutativity. 
The results are aggregated over $100$ randomly generated automata for each number of states.
The number of well-matched words used as substitutions for the variable in the SRS rule was limited to $40$.
We observe that, for the cancellation property ($c X r \to \epsilon$), increasing
the number of well-matched words for substitution does not improve results; the substitution $X \mapsto \epsilon$
is sufficient to achieve the same effect. 
However, this is not true for reliability, which is discussed below.
\input{plot.tex}

\Paragraph{The probabilistic setting}
%In this section, we evaluate two aspects: the impact of SRS on reliability and the reduction in processing time.
Since the SRS enables equivalence testing to be performed in two stages, 
we compare the algorithm with advice against the baseline, using two configurations for the latter: 
$10^5$ and $2 \cdot 10^5$ random words.
The second configuration checks whether the improvements observed with advice are genuinely due to the SRS, 
rather than a result of evaluating a larger number of random words.
The experimental setup consists of $10^4$ randomly generated automata, each with $10$ to $25$ states.
%The input alphabet was generated individually for each automaton, with slight variations depending on the considered property.
%In particular, the number of symbols of each type ranged from $2$ to $5$, while the number of stack symbols ranged from $2$ to $16$.
The number of well-matched words used for SRS was limited to $200$.

\begin{table}[H]
\centering
\begin{tabular}{cc|ccc}
\hline
SRS rule & Metric
& $10^5$ random words
& $2\cdot10^5$ random words
& $10^5$ with SRS \\
\hline
\multirow{2}{*}{Cancellation}
& Success ratio & 30.29\% & 40.1\% & 57.19\% \\
& Avg. time [s] & 0.69 & 1.28 & 2.6 \\
\hline
\multirow{2}{*}{Commutativity}
& Success ratio & 73.93\% & 78.59\% & 74.11\% \\
& Avg. time [s] & 0.69 & 0.89 & 2.0 \\
\hline
\multirow{2}{*}{Idempotency}
& Success ratio & 91.0\% & ... & 91.69\% \\
& Avg. time [s] & 0.38 & ... & 0.92 \\
\hline
\end{tabular}
\caption{Comparison of success ratio and average execution time for different SRS rule types}
\end{table}

\Paragraph{Summary}
The presented results show a reduction of approximately $20\%$ in the number of equivalence queries for commutativity and idempotency,
and around $50\%$ for cancellation rules. In our setup, the time measurements indicate that the use of SRS
generally increases the overall execution time. However, SRS reduces execution time for the largest instances.
This suggests that the benefit of the advice mechanism grows with automaton size.
Based on experiments for DFA~\cite{ecai25}, we expected the most significant gains for commutativity.
However, the obtained results differ substantially from these expectations.
We suspect the generated \DVPA consistent commutativity with were too simple for advice to help. 
%constrained limited the impact of the SRS --- the autmata were too simple. 
%this rule.
% We hypothesize that the method
%used to generate the automata did not allow the full potential of this rule to be observed.
The reliability results do not show a significant improvement.
Nevertheless, SRS improves the success ratio, while the overhead mainly comes
from an auxiliary heuristic for selecting well-matched words, intended to increase the chances of detecting inconsistency.
Its cost is negligible comparing to the exact equivalence check but becomes noticeable with the probabilistic one.
%If the evaluation of words becomes more expensive (e.g. LLM as a membership oracle), the overhead of our method vanishes. 
This suggests that SRS can be considered a useful feature in a heuristic approach.


% %\begin{itemize}
%     \item For \emph{well-matched commutativity}, we generate two \DVPA over disjoint alphabets, take the product of these automata, where
% undefined transitions are defined as self-loops. 
%     \item For \emph{well-matched cancellation}, we generate a \DVPA $\aut$ randomly over the alphabet without letter $c,r$. 
%     Then, $\aut$ is extended to $\aut'$ over the alphabet including $c,r$ such that the rule $c X r \to \epsilon$ is consistent.   
%     \item For \emph{well-matched idempotency}, we generate a SEVPA with the condition that returning to 
%     the same module does not change the state, which implies that $c c X r r \to c  X r$ is consistent.
% \end{itemize}


