In the previous section, 
we have studied how advice provided via an \SRS reduces the number of equivalence queries necessary to learn an unknown language.
In this section, we address the converse: \emph{synthesis} of rewrite rules consistent with a given visibly pushdown language. 
Synthesized rewrite rules can serve as an explanation of the visibly pushdown language under consideration, capturing its key properties through concise rewrite rules.

There are two possible variants of this problem:
\begin{description}
\item[\textit{Synthesis from Automata}:] Given a \DVPA, generate string rewrite rules consistent with the language of the given \DVPA.
\item[\textit{Synthesis from Oracles}:] Given an oracle for an unknown visibly pushdown language, generate string rewrite rules consistent with the language. 
\end{description}

While \textit{Synthesis from Oracles} seems more natural, we show that even given a \DVPA, the synthesis problem is challenging for two reasons: 
high computational complexity and the lack of a clear objective --- it is not obvious what constitutes a ``relevant'' rule. 
Therefore, \textit{Synthesis from Oracles} suffers from these same shortcomings, and further progress in \textit{Synthesis from Automata} is required before \textit{Synthesis from Oracles} can be effectively addressed.
Consequently, for the remainder of this section we focus on \textit{Synthesis from Automata}.



We establish a polynomial upper bound on the lengths of words in minimal-length rules consistent with $\aut$, which implies that such rules
can be computed in non-deterministic polynomial time. 
In order to achieve that, we give  sufficient conditions for a pair of words to form a rewrite rule consistent with $\lang(\aut)$.
We consider first rules without variables, and then move to rules with variables.

We consider a \DVPA $\aut$ with the set of states $Q$ over a non-trivial alphabet, i.e., either there are at least two local letters or at least one pair of a call and a return letter.
Under these assumptions, there are exponentially many well-matched words.

\Paragraph{The length of a minimal consistent well-matched rule: the ground case}
Consider a well-matched word $w$. We associate with this word a function $f_w \colon Q \to Q$.
Such functions correspond to equivalence classes of the syntactic congruence of a \DFA.
Formally, for a state $q \in Q$ we define $f_w(q) = q'$ if and only if for some stack content $u$, we have $(q,u) \trans{\aut}{a} (q',u)$.
Note that the existential quantifier over $u$ can be replaced by a universal one, since transitions over well-matched words are stack-independent. 
If $w_1, w_2$ are well-matched words such that $f_{w_1} = f_{w_2}$, 
then for all words $x,y$ we have $x w_1 y \in \lang(\aut) \iff x w_2 y \in \lang(\aut)$~\cite[Theorem~1]{alur2005congruences}.
It follows that $w_1 \to w_2$ is consistent with $\lang(\aut)$. 
Setting $n = |Q|$, there are $n^n$ functions from $Q$ to $Q$, and hence there are $n^n = 2^{\bigO(n \log n)}$ distinct functions $f_w$ over all well-matched words $w$.
There are $2^{\Theta(k)}$ well-matched words of the length $k$, and hence among well-matched words of length $\bigO(n \log n)$ 
there is a pair of well-matched words $\lhs, \rhs$ such that $f_{\lhs} = f_{\rhs}$. 
It follows that $\lhs \to \rhs$ is a ground well-matched rule consistent with $\lang(\aut)$. 

\Paragraph{The length of a minimal consistent well-matched rule: the general case}
For rules with variables, we consider only a subset of rules and only a sufficient condition for consistency. 
Nevertheless, we obtain asymptotically the same bound. 
Consider a rewrite rule with variables $\lhs \to \rhs$ of the form:
\begin{align*}
 && w_0 X_1 w_1 X_2 \ldots  X_k w_{k} &\to v_0 X_1 v_1 X_2 \ldots  X_k v_{k}
\end{align*}
where the order of variables in $\lhs$ and $\rhs$ is the same and all words $w_i,v_i$ are well-matched words over $\vAlph$, i.e.,
do not contain variables.
Observe that if the corresponding words $w_i$ and $v_i$ define the same function, i.e., 
$f_{w_0} = f_{v_0}, \ldots, f_{w_{k}} = f_{v_k}$, then the rewrite rule $\lhs \to \rhs$ is consistent with the language of $\aut$. 
To see this, consider an instantiation of $\lhs \to \rhs$, i.e., 
consider a sequence of well-matched words $b_1, \ldots, b_k$, which we substitute for $X_1, \ldots, X_k$
obtaining the ground rewrite rule:
\begin{align*}
&&	w_0 b_1 w_1 b_2 \ldots  b_k w_{k} &\to v_0 b_1 v_1 b_2 \ldots  b_k v_{k}
\end{align*}
Observe that the left hand side and the right hand side define the same function, and hence this ground rule is 
consistent with $\lang(\aut)$.
It follows that every rule from $\sem{\lhs \to \rhs}$ is consistent with $\lang(\aut)$, which by definition means that 
$\lhs \to \rhs$ is consistent with $\lang(\aut)$.
Now, setting $n = |Q|$, there are $n^{k\cdot n}$ different combinations of $k$ functions from $Q$ to $Q$. 
The number of well-matched words is $2^{\Omega(n)}$ and hence there is a constant $c>0$ such that
among words of the form $w_0 X_1 w_1 X_2 \ldots  X_k w_{k}$ such that 
the sum of lengths of $w_0, \ldots, w_k$ is bounded by  ${c \cdot k\cdot n \cdot \log(n)}$, 
there is a pair of words that forms a rewriting rule consistent with $\lang(\aut)$.

%\Paragraph{Rule synthesis in non-deterministic polynomial time}
The counting argument shows that for a given $\DVPA$ $\aut$ there are short (of polynomial length in $|\aut|$) 
rewrite rules that are consistent with $\lang(\aut)$. 
Since testing whether a given rule (with a bounded number of variables)
is consistent with $\lang(\aut)$ is decidable in polynomial time, we can find (some) rules consistent with $\lang(\aut)$ in non-deterministic polynomial time. 

\Paragraph{Rules of exponential length}
While minimal-length rules are of polynomial length, some consistent rules are of exponential length even if we consider ``nonredundant'' rules. 
Even in the \DFA case, consider automata $\aut_n$ such that $|Q| = n$  and for all $a \in \Sigma$, the function $f_a$ is the same permutation of states of an exponential order, e.g., the composition 
of disjoint cycles of prime orders $p_1, \ldots, p_k$ has the order $p_1 \cdot \ldots \cdot p_k$.
Then, there exist words $w \neq \epsilon$ such that $w \to \epsilon$ is a rule consistent with $\lang(\aut_n)$, but every such word $w$ has length at least exponential in $n$.

While we have shown that rules can be generated in non-deterministic polynomial time, 
a natural question is whether this is optimal.

\Paragraph{Complexity of finding consistent rules}
We have established an upper bound on the minimal length of words defining the same function. 
We show that computing the exact minimal length of such words is difficult, which implies that
finding a minimal-length rewriting rule, which is consistent with a given \DVPA is difficult as well.
Furthermore, the problem is already difficult  for \DFA.

We consider the \emph{minimal collision} problem defined as follows: given a \DFA $\aut$ and $k>0$, 
does there exist a pair of different words $w_1, w_2$ such that (1)~$|w_1|,|w_2| \leq k$, and
(2)~$f_{w_1} = f_{w_2}$. As we have discussed above for the ground rewrite rules, if $|\Sigma| \geq 2$, then for every $k > |\aut|\cdot \log(|\aut|)$ the answer is yes. 
Consequently, the problem is in \NP. We show that it is \NP-complete.

\begin{restatable}{lemma}{NPHardnessOfClash}
The minimal collision problem is \NP-complete.
\end{restatable}
\begin{proof}[Proof's sketch]
The proof is by reduction from the 3-SAT problem.
The main difficulty is to ensure that if the formula is unsatisfiable, 
then no pair of short words defines the same function from $Q$ to $Q$ .
We construct a gadget \DFA such that unless one of the designed letters occurs in a word, 
each word of length at most $n$ generates a unique function.
We consider the alphabet consisting of literals from a given 3-CNF formula.
With that gadget \DFA, we enforce that short words having a collision must contain a variable or its negation, i.e., 
represent an assignment, and these words must contain a literal from each clause.
\end{proof}


% In the minimal collision problem, both sides of the rule are unknown words.
% We discuss another decision problem associated with rule synthesis, in which the right-hand side of the rule is fixed. 
% For simplicity, we consider the right-hand side to be a local letter.
% The \emph{bounded rewriting to single letter} problem asks, given a \DFA $\aut$ over $\Sigma$,  a letter $a \in \Sigma$, and a bound $k>0$, 
% whether there exists a word $w \in (\Sigma \setminus \set{a})^*$ such that 
% (1)~$w \to a$ is a rule consistent with $\lang(\aut)$, and
% (2)~$|w| \leq k$. 
% This problem is closely related to the \emph{membership}  problem for monoids~\cite{Kozen77} and groups~\cite{DBLP:journals/tcs/Jerrum85}.
% The membership problem for monoids reduces to the \emph{bounded automata intersection} problem~\cite{Kozen77,DBLP:conf/wia/Wareham00}, which 
% asks, given $m,k>0$ and \DFA $\aut_1, \ldots, \aut_m$, whether there exists a word $w$ such that $|w| \leq k$ and $w \in \lang(\aut_i)$ for every $i$.
% The bounded automata intersection problem is \PSPACE-complete if $k$ is given in binary~\cite{Kozen77} and \NP-complete if $k$ is given in unary~\cite{DBLP:conf/wia/Wareham00}.
% The same complexity bounds apply to the bounded rewriting to single letter problem as there is a straightforward reduction from the bounded automata intersection problem.
% Clearly, the problem is in $\PSPACE$ for $k$ given in binary and $\NP$ for $k$ given in unary.

% \begin{restatable}{lemma}{ReductionToLocal}
% The bounded rewriting to a single letter with the parameter $k$ is 
% (a)~\PSPACE-complete, if $k$ is given in binary, and
% (b)~\NP-complete, if $k$ is given in unary.
% \end{restatable}

In summary, there is no algorithm for generating minimal-length consistent rules in deterministic polynomial time unless \PTIME = \NP.
%\todo{ soften it slightly}
%Consequently, further research on the rule synthesis problem should focus on efficient heuristics rather than exact algorithms.

%% \cite{DBLP:journals/aaecc/ThiemannZGS08}
%% Differences between TRS and SRS and why synchornzing words do not reduce to SRS rule synthesis
%% with TRS we can express that a rule correwsponds to a reset word. It is not possible with an SRS.
%% 
